Quasi-universal acceptance of basic science in the US

### 2024-09-12 appendix_acceptance_knowledge.Rmd
first version within word limit of 4k

---
title             : "Quasi-universal acceptance of basic science in the US"
shorttitle        : "Does anyone mistrust basic science?"

header-includes:  | # to prevent floats from moving past certain points (for the appendix)
  \usepackage{placeins}

author: 
  - name          : ""
    affiliation   : ""

affiliation:
  - id            : ""
    institution   : ""

abstract: |
  XX

keywords          :

floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output            : papaja::apa6_doc # "_doc" for word; however, note that some of the features of kableExtra are not available for word and will yield errors. For example to knit to word, you'll have to comment out all "add_header_above()" functions for model output tables, or e.g. set always allow html as true in the yaml heading

always_allow_html: true

appendix:
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"

bibliography: references.bib
---

```{r setup, include=FALSE}
# Figure out output format
is_docx <- knitr::pandoc_to("docx") | knitr::pandoc_to("odt")
is_latex <- knitr::pandoc_to("latex")
is_html <- knitr::pandoc_to("html")

# Word-specific things
table_format <- ifelse(is_docx, "huxtable", "kableExtra")  # Huxtable tables
conditional_dpi <- ifelse(is_docx, 300, 300)  # Higher DPI
conditional_align <- ifelse(is_docx, "default", "center")  # Word doesn't support align

# Knitr options
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  # tidy.opts = list(width.cutoff = 120),  # Code width
  # fig.retina = 3, dpi = conditional_dpi,
  # fig.width = 7, fig.asp = 0.618,
  # fig.align = conditional_align, out.width = "100%",
  fig.path = "output/figures/",
  cache.path = "output/_cache/",
  fig.process = function(x) {  # Remove "-1" from figure names
    x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
    if (file.rename(x, x2)) x2 else x
  },
  options(scipen = 99999999)  # Prevent scientific notation
)

# R options
options(
  width = 90,  # Output width
  dplyr.summarise.inform = FALSE,  # Turn off dplyr's summarize() auto messages
  knitr.kable.NA = "",  # Make NAs blank in kables
  kableExtra.latex.load_packages = FALSE,  # Don't add LaTeX preamble stuff
  modelsummary_factory_default = table_format,  # Set modelsummary backend
  modelsummary_format_numeric_latex = "plain"  # Don't use siunitx
)
```

```{r packages, include=FALSE}
# load required packages
library("papaja")      # For APA style manuscript   
library("lme4")        # model specification / estimation
library("lmerTest")    # provides p-values in the output
library("tidyverse")   # data wrangling and visualisation
library("afex")        # anova and deriving p-values from lmer
library("broom")       # extracting data from model fits 
library("broom.mixed") # extracting data from mixed models
library("metafor")     # doing mata analysis
library("patchwork")   # put several plots together
library("ggridges")    # for plots
library("gghalves")    # for plots
library("ggbeeswarm")  # Special distribution-shaped point jittering
library("knitr")       # for tables
library("kableExtra")  # also for tables
library("ggpubr")      # for combining plots with ggarrange() 
library("grid")        # for image plots   
library("gridExtra")   # for image plots
library("png")         # for image plots
library("modelsummary") # for regression tables
library("ggrepel") # for better random patterns in plots
```

```{r functions}
# load plot theme
source("functions/plot_theme.R")

# load other functions
source("functions/own_functions.R")
```

# Introduction

Trust in science is related to many desirable outcomes, from acceptance of anthropogenic climate change (ref) or vaccination (ref) to following recommendations during COVID [@alganTrustScientistsTimes2021], which suggests that trust in science was the most important predictor of these behaviors).

Unfortunately, people who report a high degree of trust in science are a minority in most countries, and they are outnumbered by people who have low trust in science in many areas, e.g., most of Africa and significant parts of Asia [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Moreover, trust in science has recently been declining in some countries [@alganTrustScientistsTimes2021; although see @wellcomeglobalmonitorPublicTrustScientists2021], including the US [@brianAmericansTrustScientists2023], where it is also increasingly polarizing [@gauchatPoliticizationSciencePublic2012; @krauseTrendsAmericansTrust2019a; @liPolarizationPublicTrust2022]. Rejection of science can take the more extreme form of conspiracy theories, many of which question the scientific consensus on vaccination, climate change, and even the shape of the Earth.

What does this apparent rejection of science actually entail? Do people who say they do not trust science, or who believe in conspiracy theories at odds with well-established science, reject most of science? Or, on the contrary, do they object to a few specific facets of science, while still accepting the overwhelming majority of basic science? Answering this question has theoretical implications: if some people reject science wholesale, then it’s possible that their distrust of science causes their rejection of specific scientific theories, or facilitates their acceptance of conspiracy theories; by contrast, if everyone trusts most of science, then the mistrust of specific facets of science is more likely to be post-hoc, a rationalization of other beliefs or behaviors. The present question also has practical implications: many communication attempts leverage the scientific consensus [e.g. on vaccination, climate change, etc., for review, see @vanstekelenburgScientificConsensusCommunicationContested2022; see also @veckalov27countryTestCommunicating2024]. These attempts are more likely to be successful if everyone trusts basic science than if some people reject science wholesale.

Studies of science knowledge provide relevant evidence, as people who answer science knowledge questions correctly presumably trust scientists in this respect. The average level of science knowledge is quite low, even in countries with extensive science education. For example, only 55% in the US (in 2014) and only 46% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses [@committeeonscienceliteracyandpublicperceptionofscienceScienceLiteracyConcepts2016]. However, these results only provide a lower bound on how much people trust basic science, as participants who do not provide the correct answer might accept it when it is pointed out to them.

This relative lack of trust in science could mean different things. It could mean that some people reject science wholesale, or at least to a significant extent. Or it could mean that they maintain a high degree of trust in most of science, and only question specific results that happen to contradict specific other beliefs they have (e.g. creationism), or beliefs that are associated with disliked behaviors (e.g. vaccination, acting for climate change).

We answer this question by looking at how much Americans accept basic scientific knowledge. We pay special attention to two groups of participants: people who say they don’t trust science (much, or at all), and people who endorse science-related conspiracy theories (CTs), or have a CT mindset, since this has been linked with a low degree of trust in science, and that many CTs imply a mistrust of at least some science (vaccination, hydroxychloroquine, etc.).

We briefly review work on trust in science, and science knowledge, before turning to work on CT and trust in science.

## Trust in science

Across the globe, most people say they trust science at least to some extent. In 2018, the Wellcome Global Monitor (WGM) surveyed of over 140'000 people in over 140 countries on trust in science [@wellcomeglobalmonitorWellcomeGlobalMonitor2018]. In 2020, during the first year of the Covid pandemic, a follow up survey was made in 113 countries, involving 119'000 participants [@wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Across all countries, in 2018, 32% of participants said they trust science "a lot" (41% in 2020), 45% trust "some" science (39% in 2020), and only 13% percent trust science "not much/ not at all" (13% in 2020), while 10% indicated "don't know" (7% in 2020)[^1].

[^1]: <https://wellcome.org/reports/wellcome-global-monitor-covid-19/2020#gid=6acd&pid=0>\>

Trust in science also appears to be relatively stable in the US [@funkPublicConfidenceScientists2020]. However, there are some potentially worrying trends: Recently, trust in US has dropped [@brianAmericansTrustScientists2023], and some evidence suggests increasing polarization of trust in science along political lines [@gauchatPoliticizationSciencePublic2012]. For example, the partisan divide on the question of whether climate change is a major threat to society has been growing in the last decade[^2] [@kennedyWhatDataSays2023]. A recent study suggests that by some measure, only about half of the US population believes in anthropogenic climate change, and that individuals do mostly not seem to change their mind on on the issue [@mottaChangingMindsChanging2021].

[^2]: <https://www.pewresearch.org/short-reads/2023/08/09/what-the-data-says-about-americans-views-of-climate-change/>

Trust in science is typically measured in three ways: either asking explicitly about general trust in science or scientists, by asking about science attitudes (e.g. support for public funding of science), or asking questions about specific, typically contentious, science topics (e.g. vaccines, climate change, GMOs). It is unclear to which extent these measures capture trust in basic science. To assess this, we can look at the literature on science knowledge.

[introduce SK as: that gives us a lower bound of how much people trust basic science]

## Science knowledge

Much of the research on public understanding of science measures to which extent people have basic science knowledge (e.g. "Does the Earth go around the Sun, or does the Sun go around the Earth?"). For us, these surveys provide a lower bound to how much people trust basic science: people who give the right answer (and not by chance) supposedly trust science on that question. Unfortunately, that lower bound has often been indeed pretty low: researchers have pointed out the deficit in basic science knowledge [@millerPublicUnderstandingAttitudes2004; @millerMeasurementCivicScientific1998a]. For example, only 55% in the US (in 2014) and only 46% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses [@committeeonscienceliteracyandpublicperceptionofscienceScienceLiteracyConcepts2016]. A question that remains is whether people do not provide the right answers because they are not familiar with the scientific consensus, or because they explicitly reject that consensus.

[To do: Make a descriptive overview of science knowledge across time/countries.]

## CT and TiS

Recently, the relationship of one group to trust in science (TiS) has been particularly scrutinized: conspiracy theorists (CTists), either defined as people who endorse specific conspiracy theories, or as people who have a conspiracy mindset [@rutjensConspiracyBeliefsScience2022; @vranicDidMyOwn2022]. CTists reject specific science knowledge (related to their CT), and they tend to trust science less on the whole. What remains unclear is to which extent CTists (dis)trust basic science.

## The present studies

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted correct answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants’ trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (OSF)). The differences between the four studies are summarized presently, and the methods are detailed below.

### Materials

In Study 1, we used questions drawn from questionnaires of scientific knowledge (e.g. "Are electrons smaller, larger, or the same size as atoms? [Smaller; Same size; Larger]"), supplemented by a ‘trick’ question ("Where do trees mainly draw the materials with which they create their mass? [Earth; Water; Air]"; correct answer: Air). In Studies 2 and 3, this last question was removed. The scientific facts used in Studies 1 to 3 represent long-established and basic knowledge. In Study 4 we used more recent much less basic scientific discoveries (e.g. "How many more glial cells are there in the brain in comparison with neurons, as established in 2016? [The same amount; Twice as many; Tenth as many]"; correct answer: The same amount).

### Presentation of the scientific consensus

In Study 1, we simply told participants that they would be provided with the scientifically consensual answer. However, for participants to accept this answer, they must not only trust science, but also trust us (that we are presenting them with the actual scientifically consensual answer). To remove this issue, in Studies 2 to 3, we presented participants with a short explanation of the correct answer, as well as links to three sources per answer (e.g. Wikipedia, National Geographic or NASA). In Study 4, as the topics were more complex, we did not provide an explanation, but we provided two sources per answer.

### Measure of acceptance of the scientific consensus

In Study 1, we simply look at whether participants say that they accept the scientifically consensual answer. In the subsequent studies, we asked participants to explain why they disagreed with the scientific consensus. This revealed that a number of participants had made a mistake (misunderstanding, selecting the wrong answer). As a result, in Studies 3 and 4, participants who have indicated that they rejected the scientifically consensual answer were offered the option to revise their answer, or to keep rejecting it.

### Additional questions

In Studies 3 and 4, we attempted to understand how some people who say they do not trust science still accept scientifically consensual answers, by asking them whether they accepted the answers on the basis of trust in science or because they had independently verified them.

### Samples

Studies 1 and 2 were conducted on the standard sample of US participants recruited on the platform Prolific Academic. In order to increase the share of participants with low trust in science, and who endorse conspiracy theories, Studies 3 and 4 used the same platform, but only recruited participants who had declared previously being skeptical of vaccination.

### Hypotheses

The main goal of the present studies is descriptive: to find out whether participants who report not trusting science, or who believe in conspiracy theories, still accept most well-established scientific facts. However, we also tested two directional hypotheses (pre-registered as research questions in the Study 1):

**H1: Higher trust in science is associated with more science knowledge and more acceptance of the scientific consensus**

**H2: Higher conspiracy thinking/belief is associated with less science knowledge and less acceptance of the scientific consensus**

# Methods

```{r exp1}
# Analyze data of experiments and store results

# Experiment 1

# read the cleaned long version data set
exp1_long <- read_csv("exp_1/data/cleaned_long.csv")

# read wide version data set
exp1_wide <- read_csv("exp_1/data/cleaned_wide.csv")

# Research questions

# RQ3
cor_trust_knowledge <- cor.test(exp1_wide$wgm_sciencegeneral, exp1_wide$avg_knowledge) %>% text_ready()

cor_trust_acceptance <- cor.test(exp1_wide$wgm_sciencegeneral, exp1_wide$avg_acceptance) %>% text_ready()

# RQ4
cor_conspiracy_knowledge <- cor.test(exp1_wide$BCTI_avg, exp1_wide$avg_knowledge) %>% text_ready()

cor_conspiracy_acceptance <- cor.test(exp1_wide$BCTI_avg, exp1_wide$avg_acceptance)%>% text_ready()

# Exploratory: Acceptance and trust in science/conspiracy thinking conditional on false answers only
exp1_false_knowledge <- exp1_long %>% 
  # make numeric versions
  mutate(acceptance_num = ifelse(acceptance == "Yes", 1, 0)
  ) %>% 
  group_by(id, knowledge) %>% 
  # calculate by-participant averages
  summarize(
    n = n(),
    n_accepted = sum(acceptance_num),
    avg_acceptance  = sum(acceptance_num)/n(), 
    wgm_sciencegeneral = mean(wgm_sciencegeneral), 
    BCTI_avg = mean(BCTI_avg), 
    # add additonal measures for robustness checks
    CMQ_avg = mean(CMQ_avg), 
    SICBS = mean(SICBS),
    wgm_scientists = mean(wgm_scientists), 
    pew = mean(pew)
  ) %>% 
  # filter to only false responses
  filter(knowledge == FALSE) %>% 
  ungroup()

false_answers_cor_trust_acceptance <- cor.test(exp1_false_knowledge$wgm_sciencegeneral, exp1_false_knowledge$avg_acceptance) %>% text_ready()

false_answers_cor_conspiracy_acceptance <- cor.test(exp1_false_knowledge$BCTI_avg, exp1_false_knowledge$avg_acceptance) %>% text_ready()

# extract descriptives for inline reporting
exp1_descriptives <- list(
  # Demographics
  n_subj = n_distinct(exp1_wide$id),
  gender = exp1_wide %>% group_by(gender) %>% summarize(n = n_distinct(id)) %>% split(.$gender),
  age = exp1_wide %>% summarize(across(age,list(mean = mean, median = median, sd = sd), 
                                       .names = "{.fn}")) %>% rounded_numbers(),
  # RQ1 & RQ2
  means = exp1_wide %>% 
    summarize(across(c(avg_knowledge, avg_acceptance), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()) %>% 
    mutate_if(is.numeric, round, digits = 2) %>% 
    mutate(across(ends_with("_mean"), ~paste0(.*100, " %"))),
  means_numeric = exp1_wide %>% 
    summarize(across(c(avg_knowledge, avg_acceptance,), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()),
  # RQ3
  cor_trust_knowledge = cor_trust_knowledge,
  cor_trust_acceptance = cor_trust_acceptance,
  # RQ4
  cor_conspiracy_knowledge = cor_conspiracy_knowledge,
  cor_conspiracy_acceptance = cor_conspiracy_acceptance, 
  # Exploratory
  # acceptance and knowledge
  conditional_acceptance = exp1_long %>% 
    group_by(knowledge, acceptance) %>% 
    count() %>% 
    group_by(knowledge) %>% 
    mutate (share = n/sum(n), 
            # rename knowledge values
            knowledge = ifelse(knowledge == FALSE, "false", "true")) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate (share = paste0(share*100, " %")) %>% 
    super_split(knowledge, acceptance),
  # acceptance and trust in science/conspiracy thinking
  false_answers_cor_trust_acceptance = false_answers_cor_trust_acceptance,
  false_answers_cor_conspiracy_acceptance = false_answers_cor_conspiracy_acceptance
)
```

```{r exp2}
# Analyze data of experiments and store results

# Experiment 2

# read the cleaned long version data set
exp2_long <- read_csv("exp_2/data/cleaned_long.csv")

# read wide version data set
exp2_wide <- read_csv("exp_2/data/cleaned_wide.csv")

# read coded justifications
exp2_justifications <- read_csv("exp_2/data/justifications_clean.csv")

# Research questions

# H1a
cor_trust_knowledge <- cor.test(exp2_wide$wgm_sciencegeneral, exp2_wide$avg_knowledge) %>% text_ready()

# H2a
cor_conspiracy_knowledge <- cor.test(exp2_wide$BCTI_avg, exp2_wide$avg_knowledge) %>% text_ready()

# Exploratory: Acceptance and trust in science/conspiracy thinking conditional on false answers only
exp2_false_knowledge <- exp2_long %>% 
  # make numeric versions
  mutate(acceptance_num = ifelse(acceptance == "Yes", 1, 0)
  ) %>% 
  group_by(id, knowledge) %>% 
  # calculate by-participant averages
  summarize(
    n = n(),
    n_accepted = sum(acceptance_num),
    avg_acceptance  = sum(acceptance_num)/n(), 
    wgm_sciencegeneral = mean(wgm_sciencegeneral), 
    BCTI_avg = mean(BCTI_avg), 
    # add additonal measures for robustness checks
    CMQ_avg = mean(CMQ_avg), 
    SICBS = mean(SICBS),
    wgm_scientists = mean(wgm_scientists), 
    pew = mean(pew)
  ) %>% 
  # filter to only false responses
  filter(knowledge == FALSE) %>% 
  ungroup()

# H1b
# conditional on false answers
false_answers_cor_trust_acceptance <- cor.test(exp2_false_knowledge$wgm_sciencegeneral, exp2_false_knowledge$avg_acceptance) %>% text_ready()
# pooled across all cases (in line with all other studies' analysis)
cor_trust_acceptance <- cor.test(exp2_wide$wgm_sciencegeneral, exp2_wide$avg_acceptance) %>% text_ready()

# H2b
# conditional on false answers
false_answers_cor_conspiracy_acceptance <- cor.test(exp2_false_knowledge$BCTI_avg, exp2_false_knowledge$avg_acceptance) %>% text_ready()
# pooled across all cases (in line with all other studies' analysis)
cor_conspiracy_acceptance <- cor.test(exp2_wide$BCTI_avg, exp2_wide$avg_acceptance) %>% text_ready()

# extract descriptives for inline reporting
exp2_descriptives <- list(
  # Demographics
  n_subj = n_distinct(exp2_wide$id),
  gender = exp2_wide %>% group_by(gender) %>% summarize(n = n_distinct(id)) %>% split(.$gender),
  age = exp2_wide %>% summarize(across(age,list(mean = mean, median = median, sd = sd), 
                                       .names = "{.fn}")) %>% rounded_numbers(),
  # RQ1 & RQ2
  means = exp2_wide %>% 
    summarize(across(c(avg_knowledge, avg_acceptance), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()) %>% 
    mutate_if(is.numeric, round, digits = 2) %>% 
    mutate(across(ends_with("_mean"), ~paste0(.*100, " %"))) ,
  means_numeric = exp2_wide %>% 
    summarize(across(c(avg_knowledge, avg_acceptance,), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()),
  # H1
  # a
  cor_trust_knowledge = cor_trust_knowledge,
  # b (conditional on false responses)
  false_answers_cor_trust_acceptance = false_answers_cor_trust_acceptance,
  # b (all cases)
  cor_trust_acceptance = cor_trust_acceptance,
  # H2
  # a
  cor_conspiracy_knowledge = cor_conspiracy_knowledge,
  # b (conditional on false responses)
  false_answers_cor_conspiracy_acceptance = false_answers_cor_conspiracy_acceptance,
  # b(all cases)
  cor_conspiracy_acceptance = cor_conspiracy_acceptance, 
  # Exploratory
  # acceptance and knowledge
  conditional_acceptance = exp2_long %>% 
    group_by(knowledge, acceptance) %>% 
    count() %>% 
    group_by(knowledge) %>% 
    mutate (share = n/sum(n), 
            # rename knowledge values
            knowledge = ifelse(knowledge == FALSE, "false", "true")) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate (share = paste0(share*100, " %")) %>% 
    super_split(knowledge, acceptance),
  # Justifications
  justifications_n = nrow(exp2_justifications), 
  justifications_n_participants = n_distinct(exp2_justifications$id),
  justification_by_category = exp2_justifications %>%
    group_by(category)%>%
    summarize(n = n(), 
              n_subjects = n_distinct(id)) %>%
    mutate(share = n/sum(n)) %>%
    mutate_if(is.numeric, round, digits=3) %>% 
    mutate(share = paste0(share*100, "%")) %>% 
    split(.$category)
)
```

```{r exp3}
# Analyze data of experiments and store results

# Experiment 3

# read the cleaned long version data set
exp3_long <- read_csv("exp_3/data/cleaned_long.csv")

# read wide version data set
exp3_wide <- read_csv("exp_3/data/cleaned_wide.csv")

# read coded justifications
exp3_justifications <- read_csv("exp_3/data/justifications_clean.csv")

# Research questions

# H1
cor_trust_knowledge <- cor.test(exp3_wide$wgm_sciencegeneral, exp3_wide$avg_knowledge,
  use = "complete.obs") %>% text_ready()
cor_trust_acceptance <- cor.test(exp3_wide$wgm_sciencegeneral, exp3_wide$avg_acceptance,
  use = "complete.obs") %>% text_ready()

# H2
cor_conspiracy_knowledge <- cor.test(exp3_wide$BCTI_avg, exp3_wide$avg_knowledge,
  use = "complete.obs") %>% text_ready()
cor_conspiracy_acceptance <- cor.test(exp3_wide$BCTI_avg, exp3_wide$avg_acceptance,
  use = "complete.obs") %>% text_ready()

# Exploratory: Acceptance and trust in science/conspiracy thinking conditional on false answers only
exp3_false_knowledge <- exp3_long %>% 
  # make numeric versions
  mutate(acceptance_num = ifelse(acceptance == "Yes", 1, 0)
  ) %>% 
  group_by(id, knowledge) %>% 
  # calculate by-participant averages
  summarize(
    n = n(),
    n_accepted = sum(acceptance_num),
    avg_acceptance  = sum(acceptance_num)/n(), 
    wgm_sciencegeneral = mean(wgm_sciencegeneral), 
    BCTI_avg = mean(BCTI_avg), 
    # add additonal measures for robustness checks
    CMQ_avg = mean(CMQ_avg), 
    SICBS = mean(SICBS),
    wgm_scientists = mean(wgm_scientists), 
    pew = mean(pew)
  ) %>% 
  # filter to only false responses
  filter(knowledge == FALSE) %>% 
  ungroup()

# H1b conditional on false answers
false_answers_cor_trust_acceptance <- cor.test(exp3_false_knowledge$wgm_sciencegeneral, exp3_false_knowledge$avg_acceptance,
  use = "complete.obs") %>% 
  text_ready()

# H2b conditional on false answers
false_answers_cor_conspiracy_acceptance <- cor.test(exp3_false_knowledge$BCTI_avg, exp3_false_knowledge$avg_acceptance,
  use = "complete.obs") %>% 
  text_ready()

# Exploratory: Differences by reason for trust
trust_by_reason_model <- run_regression(exp3_wide, dependent_var = "wgm_sciencegeneral", independent_var = "reason_agreement")

acceptance_by_reason_model <-run_regression(exp3_wide, dependent_var = "avg_acceptance", independent_var = "reason_agreement")

# extract descriptives for inline reporting
exp3_descriptives <- list(
  # Demographics
  n_subj = n_distinct(exp3_wide$id),
  n_subject_by_outcome = exp3_wide %>% 
    summarize(across(c("wgm_sciencegeneral", "reason_agreement", "BCTI_avg"), 
                     ~sum(!is.na(.x)
                     )
    )
    ),
  gender = exp3_wide %>% group_by(gender) %>% summarize(n = n_distinct(id)) %>% split(.$gender),
  age = exp3_wide %>% summarize(across(age,list(mean = mean, median = median, sd = sd), 
                                       .names = "{.fn}")) %>% rounded_numbers(),
  # RQ1 & RQ2
  means = exp3_wide %>% 
    summarize(across(c(avg_knowledge, avg_acceptance, avg_acceptance_initial), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()) %>% 
    mutate_if(is.numeric, round, digits = 2) %>% 
  mutate(across(ends_with("_mean"), ~paste0(.*100, " %"))) ,
  means_numeric = exp3_wide %>% 
    summarize(across(c(avg_knowledge, avg_acceptance, avg_acceptance_initial), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()),
  # H1
  # a
  cor_trust_knowledge = cor_trust_knowledge,
  # b (conditional on false responses)
  false_answers_cor_trust_acceptance = false_answers_cor_trust_acceptance,
  # b (all cases)
  cor_trust_acceptance = cor_trust_acceptance,
  # H2
  # a
  cor_conspiracy_knowledge = cor_conspiracy_knowledge,
  # b (conditional on false responses)
  false_answers_cor_conspiracy_acceptance = false_answers_cor_conspiracy_acceptance,
  # b(all cases)
  cor_conspiracy_acceptance = cor_conspiracy_acceptance, 
  # Exploratory
  # acceptance and knowledge
  conditional_acceptance = exp3_long %>% 
    pivot_longer(c(acceptance, acceptance_initial), 
                 names_to = "measure", 
                 values_to = "acceptance") %>% 
    group_by(knowledge, measure, acceptance) %>% 
    count() %>% 
    group_by(knowledge, measure) %>% 
    mutate (share = n/sum(n), 
            # rename knowledge values
            knowledge = ifelse(knowledge == FALSE, "false", "true")) %>% 
    mutate_if(is.numeric, round, digits = 3) %>% 
    mutate (share = paste0(share*100, " %")) %>% 
    super_split(knowledge,  measure, acceptance),
  # Justifications
  justifications_n = nrow(exp3_justifications), 
  justifications_n_participants = n_distinct(exp3_justifications$id),
  justification_by_category = exp3_justifications %>%
    group_by(category)%>%
    summarize(n = n(), 
              n_subjects = n_distinct(id)) %>%
    mutate(share = n/sum(n)) %>%
    mutate_if(is.numeric, round, digits=3) %>% 
    mutate(share = paste0(share*100, "%")) %>% 
    split(.$category),
  # Consensus acceptance
  acceptance_n = exp3_wide %>% 
    filter(!is.na(reason_agreement)) %>% 
    nrow(.),
  acceptance_by_reason = exp3_wide %>% 
    group_by(reason_agreement) %>% 
    summarize(n = n()) %>%
    drop_na(reason_agreement) %>% 
    mutate(share = n/sum(n)) %>%
    mutate_if(is.numeric, round, digits=3) %>% 
    mutate(share = paste0(share*100, "%")) %>% 
    split(.$reason_agreement),
  reason_followup_n = exp3_wide %>% 
    filter(!is.na(reason_followup)) %>% 
    nrow(.),
  # Exploratory: Differences by reason for trust
  reason_means = exp3_wide %>% 
    group_by(reason_agreement) %>% 
    summarize(across(c(wgm_sciencegeneral, avg_acceptance), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()) %>% 
    mutate_if(is.numeric, round, digits = 2) %>% 
    mutate(across(acceptance_mean, ~paste0(.*100, " %"))) %>% 
    split(.$reason_agreement), 
  reason_trust_model = trust_by_reason_model %>% split(.$term), 
  reason_acceptance_model = acceptance_by_reason_model %>% split(.$term)
)
```

```{r exp4}
# Analyze data of experiments and store results

# Experiment 4

# read the cleaned long version data set
exp4_long <- read_csv("exp_4/data/cleaned_long.csv")

# read wide version data set
exp4_wide <- read_csv("exp_4/data/cleaned_wide.csv")

# read coded justifications
exp4_justifications <- read_csv("exp_4/data/justifications_clean.csv")

# Research questions

# H1
cor_trust_knowledge <- cor.test(exp4_wide$wgm_sciencegeneral, exp4_wide$avg_knowledge,
  use = "complete.obs") %>% text_ready()
cor_trust_acceptance <- cor.test(exp4_wide$wgm_sciencegeneral, exp4_wide$avg_acceptance,
  use = "complete.obs") %>% text_ready()

# H2
cor_conspiracy_knowledge <- cor.test(exp4_wide$BCTI_avg, exp4_wide$avg_knowledge,
  use = "complete.obs") %>% text_ready()
cor_conspiracy_acceptance <- cor.test(exp4_wide$BCTI_avg, exp4_wide$avg_acceptance,
  use = "complete.obs") %>% text_ready()

# Exploratory: Acceptance and trust in science/conspiracy thinking conditional on false answers only
exp4_false_knowledge <- exp4_long %>% 
  # make numeric versions
  mutate(acceptance_num = ifelse(acceptance == "Yes", 1, 0)
  ) %>% 
  group_by(id, knowledge) %>% 
  # calculate by-participant averages
  summarize(
    n = n(),
    n_accepted = sum(acceptance_num),
    avg_acceptance  = sum(acceptance_num)/n(), 
    wgm_sciencegeneral = mean(wgm_sciencegeneral), 
    BCTI_avg = mean(BCTI_avg), 
    # add additonal measures for robustness checks
    CMQ_avg = mean(CMQ_avg), 
    SICBS = mean(SICBS),
    wgm_scientists = mean(wgm_scientists), 
    pew = mean(pew)
  ) %>% 
  # filter to only false responses
  filter(knowledge == FALSE) %>% 
  ungroup()

# H1b conditional on false answers
false_answers_cor_trust_acceptance <- cor.test(exp4_false_knowledge$wgm_sciencegeneral, exp4_false_knowledge$avg_acceptance,
  use = "complete.obs") %>% 
  text_ready()

# H2b conditional on false answers
false_answers_cor_conspiracy_acceptance <- cor.test(exp4_false_knowledge$BCTI_avg, exp4_false_knowledge$avg_acceptance,
  use = "complete.obs") %>% 
  text_ready()

# Exploratory: Differences by reason for trust
trust_by_reason_model <- run_regression(exp4_wide, dependent_var = "wgm_sciencegeneral", independent_var = "reason_agreement")

acceptance_by_reason_model <- run_regression(exp4_wide, dependent_var = "avg_acceptance", independent_var = "reason_agreement")

clicks_by_reason_model <- run_regression(exp4_wide, dependent_var = "sum_link_clicks", independent_var = "reason_agreement")

# extract descriptives for inline reporting
exp4_descriptives <- list(
  # Demographics
  n_subj = n_distinct(exp4_wide$id),
  n_subject_by_outcome = exp4_wide %>% 
    summarize(across(c("wgm_sciencegeneral", "reason_agreement", "BCTI_avg"), 
                     ~sum(!is.na(.x)
                     )
    )
    ),
  gender = exp4_wide %>% group_by(gender) %>% summarize(n = n_distinct(id)) %>% split(.$gender),
  age = exp4_wide %>% summarize(across(age,list(mean = mean, median = median, sd = sd), 
                                       .names = "{.fn}")) %>% rounded_numbers(),
  # RQ1 & RQ2
  means = exp4_wide %>% 
    summarize(across(c(avg_knowledge, avg_acceptance, avg_acceptance_initial), 
                     list(mean = mean, sd = sd), 
                     .names = "{.col}_{.fn}")) %>%
    rename_with(~sub("^avg_", "", .), everything()) %>% 

>>>>

# Acceptance conditional on knowledge {#acceptance-knowledge}

Did participants who had initially gotten it wrong change their minds towards the scientific consensus? Yes. In most cases (study 1: `r exp1_descriptives$conditional_acceptance$false$Yes$share`; study 2: `r exp2_descriptives$conditional_acceptance$false$Yes$share`; study 3: `r exp3_descriptives$conditional_acceptance$false$acceptance$Yes$share`; study 4: `r exp4_descriptives$conditional_acceptance$false$acceptance$Yes$share`), participants readily accepted the scientific consensus after having initially given the wrong answer to a question. In very few cases, participants changed their mind away from the scientific consensus. These participants initially gave the correct response, but rejected the scientific consensus right after, thereby contradicting their own initial response (study 1: `r exp1_descriptives$conditional_acceptance$true$No$share`; study 2: `r exp2_descriptives$conditional_acceptance$true$No$share`; study 3: `r exp3_descriptives$conditional_acceptance$true$acceptance$No$share`; study 4: `r exp4_descriptives$conditional_acceptance$true$acceptance$No$share`). Figure \@ref(fig:acceptance-knowledge) shows acceptance rates conditional on whether the initial answer to the questions was correct or not.

(ref:acceptance-knowledge) Acceptance rates of scientific consensus, based on whether the initial response to the knowledge question was false or true.

```{r acceptance-knowledge, fig.cap="(ref:acceptance-knowledge)", fig.width=8, fig.height=8}
plot_data <- combined_data %>% 
  mutate(study = paste0("Study ", study)) %>% 
  group_by(study, knowledge, acceptance) %>% 
  count() %>% 
  group_by(study, knowledge) %>% 
  mutate (share = n/sum(n)) %>% 
  mutate_if(is.numeric, round, digits = 3)

absolute_numbers <- plot_data %>% 
  group_by(study, knowledge) %>% 
  summarise(n = sum(n))

ggplot(plot_data, aes(x = knowledge, y = share, fill = acceptance)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  geom_text(aes(label = paste0(round(share * 100, 1), "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "cividis") +
  labs(x = "Knowledge answer", fill = "Acceptance \nof consensus") +
  plot_theme +
  geom_bracket(data = absolute_numbers,
               aes(xmin = as.numeric(factor(knowledge)) - 0.4, 
                   xmax = as.numeric(factor(knowledge)) + 0.4, 
                   y.position = 1.05, label =  paste0("n = ", n)),
               inherit.aes = FALSE,
               step.increase = 0)+
  facet_wrap(~study)
```
